tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-12-04 paper.tex
Comments from Daiki

Simulation is a critical component of GWAS method development, and
generally consists of two steps:
first simulating genetic variation (genotypes)
and then simulating quantitative traits (phenotypes) based on the
synthetic genotypes.
The combined genotypes and phenotypes represent ground-truth data
which GWAS methods can be evaluated against.
Genetic variation is
usually simulated either by model-based population genetic
methods such as
\texttt{msprime}~\citep{baumdicker2022efficient}
or \texttt{SLiM}~\citep{haller2023},
or by statistical resampling from existing datasets
using methods like \texttt{HAPGEN2}~\citep{su2011hapgen2}
or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
Both approaches have advantages and disadvantages and excel
in different situations.
Roughly speaking, model-based simulation methods
provide better control of population processes such as demography,
whereas resampling methods are better at capturing
difficult to model nuances of real data.
Model-based population genetic simulations have made
great strides in recent years, with major advances in both
scalability~\citep{kelleher2016efficient,kelleher2018efficient,haller2018tree}
and realism~\citep{adrion2020,anderson2023}, and
have been successfully used to simulate large-scale GWAS
cohorts~\cite[e.g.][]{martin2017,zaidi2020}.

An important property of these population genetic simulation methods is that
they output ancestral recombination graphs (ARGs)
rather than sample genotypes.
ARGs encode the interwoven paths
of genetic inheritance arising from
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and contain rich detail about ancestral processes.
Recent breakthroughs in inferrence methods have made it possible
to estimate ARGs at biobank scale~\citep{kelleher2019,zhang2023},
and there is now intense interest in their practical
application~\citep{lewanski2023era}.
Statistical genetics has been a particular focus,
and ARG-based methods have been shown
to detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and to provide a sparse and efficient model of linkage disequilibrium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

ARG-based methods can simulate genetic variation for millions of
samples and store the output very compactly
using the ``succinct tree sequence'' encoding~\citep{wong2023general}.
and \texttt{tskit} library~\citep{ralph2020}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in gzip-compressed
VCF~\citep{danecek2011}.
At 1.36GB the original simulated ARG
(compressed using the \texttt{tszip} utility) is around 400X smaller.
Furthermore, many computations can
be expressed efficiently in terms of the underlying
ARG~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
Finally, outputting a simulated ARG provides access to the full
history, not just the genetic variation among the samples.

@@

\section{Results}
\begin{figure}[t]%
\centering
\includegraphics[width=240pt]{figures/tree-illustration.pdf}
\caption{Example simulation of a phenotype at a site with ancestral state A and
two mutations. In this diploid example each of the three individuals
is associated with two nodes (i.e., the individual with ID 0 corresponds
to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
with the null individual, $-1$. Here, the trait's causal allele is T
and effect size $\beta=0.05$. Each node in the tree has an associated
genetic value, and the overall genetic value for an individual is the
sum of their corresponding node values. The final phenotype
for each individual is the sum of the genetic value and
simulated environmental noise.
\label{fig:tree-illustration}}
\end{figure}

@@

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general an ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele and $\beta=0.05$, and
therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using familiar Python data science tools.

>>>>

Simulation is a critical component of GWAS method development, and
generally consists of two steps:
first simulating genetic variation (genotypes)
and then simulating quantitative traits (phenotypes) based on the
synthetic genotypes.
The combined genotypes and phenotypes represent ground-truth data
which GWAS methods can be evaluated against.
Genetic variation is
usually simulated either by model-based population genetic
methods such as
\texttt{msprime}~\citep{baumdicker2022efficient}
and \texttt{SLiM}~\citep{haller2023},
or by statistical resampling from existing datasets
using methods like \texttt{HAPGEN2}~\citep{su2011hapgen2}
and \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
Both approaches have advantages and disadvantages and excel
in different situations.
Roughly speaking, model-based simulation methods
provide better control of population processes such as demography,
whereas resampling methods are better at capturing
difficult to model nuances of real data.
Model-based population genetic simulations have made
great strides in recent years, with major advances in both
scalability~\citep{kelleher2016efficient,kelleher2018efficient,haller2018tree}
and realism~\citep{adrion2020,anderson2023}, and
have been successfully used to simulate large-scale GWAS
cohorts~\cite[e.g.][]{martin2017,zaidi2020}.

An important property of these population genetic simulation methods is that
they output ancestral recombination graphs (ARGs)
rather than sample genotypes.
ARGs encode the interwoven paths
of genetic inheritance arising from
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and contain rich detail about ancestral processes.
Recent breakthroughs in inferrence methods have made it possible
to estimate ARGs at biobank scale~\citep{kelleher2019,zhang2023},
and there is now intense interest in their practical
application~\citep{lewanski2023era}.
Statistical genetics has been a particular focus,
and ARG-based methods have been shown
to detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023}; to
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and to provide a sparse and efficient model of linkage disequilibrium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

ARG-based methods can simulate genetic variation for millions of
samples and store the output very compactly
in the ``succinct tree sequence'' encoding~\citep{wong2023general}
and \texttt{tskit} library~\citep{ralph2020}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in gzip-compressed
VCF~\citep{danecek2011}.
At 1.36GB the original simulated ARG
(compressed using the \texttt{tszip} utility) is around 400X smaller.
Furthermore, many calculations can
be expressed efficiently in terms of the underlying
ARG~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
Finally, outputting a simulated ARG provides access to the full
history, not just the genetic variation among the samples.

@@

\section{Results}
\begin{figure}[t]%
\centering
\includegraphics[width=240pt]{figures/tree-illustration.pdf}
\caption{Example simulation of a phenotype at a site with ancestral state A and
two mutations. In this diploid example each of the three individuals
is associated with two nodes (i.e., the individual with ID 0 corresponds
to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
with the null individual, $-1$. Here, the trait's causal allele is T
with an effect size $\beta=0.05$. Each node in the tree has an associated
genetic value, and the overall genetic value for an individual is the
sum of the genetic values of their corresponding nodes. 
The final phenotype
for each individual is the sum of the genetic value and
simulated environmental noise.
\label{fig:tree-illustration}}
\end{figure}

@@

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general an ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele with $\beta=0.05$, 
so all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using familiar Python data science tools.
